---
title: "Assignment: Local (alpha) Diversity"
author: "Mackenzie Caple; Z620: Quantitative Biodiversity, Indiana University"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW

In this exercise, we will explore aspects of local or site-specific diversity, also known as alpha ($\alpha$) diversity. 
First we will quantify two of the fundamental components of ($\alpha$) diversity: **richness** and **evenness**.
From there, we will then discuss ways to integrate richness and evenness, which will include univariate metrics of diversity along with an investigation of the **species abundance distribution (SAD)**.

## Directions:
1. In the Markdown version of this document in your cloned repo, change "Student Name" on line 3 (above) to your name.
2. Complete as much of the worksheet as possible during class.
3. Use the handout as a guide; it contains a more complete description of data sets along with the proper scripting needed to carry out the exercise.
4. Answer questions in the  worksheet.
Space for your answer is provided in this document and indicated by the ">" character.
If you need a second paragraph be sure to start the first line with ">".
You should notice that the answer is highlighted in green by RStudio (color may vary if you changed the editor theme).
5. Before you leave the classroom, **push** this file to your GitHub repo.
6. For the assignment portion of the worksheet, follow the directions at the bottom of this file. 
7. When you are done, **Knit** the text and code into a PDF file.
8. After Knitting, submit the completed exercise by creating a **pull request** via GitHub.
Your pull request should include this file `AlphaDiversity_Worskheet.Rmd` and the PDF output of `Knitr` (`AlphaDiversity_Worskheet.pdf`).


## 1) R SETUP

In the R code chunk below, please provide the code to: 
1) Clear your R environment,
2) Print your current working directory,
3) Set your working directory to your `5.AlphaDiversity` folder, and
4) Load the `vegan` R package (be sure to install first if you haven't already).

```{r}
rm(list=ls())
getwd()
setwd("~/GitHub/QB2019_Caple/2.Worksheets/5.AlphaDiversity")
require(vegan)
```

## 2) LOADING DATA

In the R code chunk below, do the following:
1) Load the BCI dataset, and 
2) Display the structure of the dataset (if the structure is long, use the `max.level = 0` argument to show the basic information).

```{r}
data(BCI)
str(BCI, max.level = 0)
```

## 3) SPECIES RICHNESS

**Species richness (S)** refers to the number of species in a system or the number of species observed in a sample.

### Observed richness

In the R code chunk below, do the following:

1. Write a function called `S.obs` to calculate observed richness

2. Use your function to determine the number of species in `site1` of the BCI data set, and

3. Compare the output of your function to the output of the `specnumber()` function in `vegan`.

```{r}
S.obs = function(x = ""){
  rowSums(x > 0) * 1
}

site1 = BCI[1, ]

S.obs(site1)
specnumber(site1)

S.obs(BCI[1:4, ])
```

***Question 1***: Does `specnumber()` from `vegan` return the same value for observed richness in `site1` as our function `S.obs`?
What is the species richness of the first four sites (i.e., rows) of the BCI matrix?

> ***Answer 1***: Yes, the two functions return the same value of observed richness.
> Site 1: 93 species
> Site 2: 84 species
> Site 3: 90 species
> Site 4: 94 species



### Coverage: How well did you sample your site?

In the R code chunk below, do the following:

1. Write a function to calculate Good's Coverage, and

2. Use that function to calculate coverage for all sites in the BCI matrix.

```{r}
C = function(x = ""){
  1 - (rowSums(x == 1) / rowSums(x))
}

C(BCI)
```
  
***Question 2***: Answer the following questions about coverage:

a.  What is the range of values that can be generated by Good's Coverage? 
b.  What would we conclude from Good's Coverage if $n_{i}$ equaled *N*?
c.  What portion of taxa in `site1` was represented by singletons? 
d.  Make some observations about coverage at the BCI plots.

> ***Answer 2a***:  From 0 to 1

> ***Answer 2b***:  Every species is only represented by one individual

> ***Answer 2c***:  7%

> ***Answer 2d***:  All sites have roughly (+/- ~4) 10% of species represented by singletons, indicating high coverage at all sites.


### Estimated richness

In the R code chunk below, do the following:

1. Load the microbial dataset (located in the `5.AlphaDiversity/data` folder),

2. Transform and transpose the data as needed (see handout),

3. Create a new vector (`soilbac1`) by indexing the bacterial OTU abundances of any site in the dataset,

4. Calculate the observed richness at that particular site, and 

5. Calculate coverage of that site

```{r}
soilbac = read.table("data/soilbac.txt", sep = "\t", header = TRUE, row.names = 1)
soilbac.t = as.data.frame(t(soilbac))
soilbac1 = soilbac.t[1,]

S.obs(soilbac1)
N = rowSums(soilbac1)
C(soilbac1)
```

***Question 3***: Answer the following questions about the soil bacterial dataset. 

a.  How many sequences did we recover from the sample `soilbac1`, i.e. *N*? 
b.  What is the observed richness of `soilbac1`? 
c.  How does coverage compare between the BCI sample (`site1`) and the KBS sample (`soilbac1`)?

> ***Answer 3a***: 2119

> ***Answer 3b***: 1074

> ***Answer 3c***: Coverage is significantly higher in the KBS sample


### Richness estimators

In the R code chunk below, do the following:

1. Write a function to calculate **Chao1**,

2. Write a function to calculate **Chao2**, 

3. Write a function to calculate **ACE**, and

4. Use these functions to estimate richness at `site1` and `soilbac1`. 

```{r}
S.chao1 = function(x = ""){
  S.obs(x) + (sum(x == 1)^2) / (2 * sum(x == 2))
}

S.chao2 = function(site = "", SbyS = ""){
  SbyS = as.data.frame(SbyS)
  x = SbyS[site, ]
  SbyS.pa = (SbyS > 0) * 1             # convert the SbyS to prescence/absence
  Q1 = sum(colSums(SbyS.pa) == 1)      # species obs. once
  Q2 = sum(colSums(SbyS.pa) == 2)      # species obs. twice
  S.chao2 = S.obs(x) + (Q1^2)/(2*Q2)
  return(S.chao2)
}

S.ace = function(x = "", thresh = 10){
  x = x[x>0]                           # excludes 0-abundance taxa
  S.abund = length(which(x > thresh))  # richness of abundant taxa
  S.rare = length(which(x <= thresh))  # richness of rare taxa
  singlt = length(which(x == 1))       # number of singleton taxa
  N.rare = sum(x[which(x <= thresh)])  # abundance of rare individuals
  C.ace = 1-(singlt / N.rare)          # coverage (prop non-singlt rare inds)
  i = c(1:thresh)                      # threshold abundance range
  count = function(i, y){
    length(y[y == i])
  }
  a.1 = sapply(i, count, x)            # number of individuals in richness i richness classes
  f.1 = (i * (i - 1)) * a.1            # k(k-1)kf sensu Gotelli
  G.ace = (S.rare / C.ace)*(sum(f.1) / (N.rare * (N.rare - 1)))
  S.ace = S.abund + (S.rare / C.ace) + (singlt / C.ace) * max(G.ace, 0)
  return(S.ace)
}

S.chao1(site1)
S.chao2(1, BCI)
S.ace(site1)

S.chao1(soilbac1)
S.chao2(1, soilbac.t)
S.ace(soilbac1)
```

***Question 4***: What is the difference between ACE and the Chao estimators? Do the estimators give consistent results? Which one would you choose to use and why?

> ***Answer 4***: The Chao estimators make estimates based on species observed once or twice within or between sites, whereas the ACE estimator uses am abundance threshold (conventionally, 10) and then creates an estimate based on this number of rare species. The estimators are reasonably consistent for the well-covered BCI data, but with the lower-coverage KBS data there are huge inconsistencies. 



### Rarefaction

In the R code chunk below, please do the following:

1. Calculate observed richness for all samples in `soilbac`,

2. Determine the size of the smallest sample,

3. Use the `rarefy()` function to rarefy each sample to this level,

4. Plot the rarefaction results, and

5. Add the 1:1 line and label.


```{r}
soilbac.S = S.obs(soilbac.t)
min.N = min(rowSums(soilbac.t))
S.rarefy = rarefy(x = soilbac.t, sample = min.N, se = TRUE)
rarecurve(x = soilbac.t, step = 20, col = "blue", cex = 0.6, las = 1)
abline(0, 1, col = 'red')
text(1500, 1500, "1:1", pos = 2, col = 'red')
```


##4) SPECIES EVENNESS
Here, we consider how abundance varies among species, that is, **species evenness**.

### Visualizing evenness: the rank abundance curve (RAC)
One of the most common ways to visualize evenness is in a **rank-abundance curve** (sometime referred to as a rank-abundance distribution or Whittaker plot).
An RAC can be constructed by ranking species from the most abundant to the least abundant without respect to species labels (and hence no worries about 'ties' in abundance). 

In the R code chunk below, do the following:

1. Write a function to construct a RAC,

2. Be sure your function removes species that have zero abundances, 

3. Order the vector (RAC) from greatest (most abundant) to least (least abundant), and 

4. Return the ranked vector

```{r}
RAC = function(x = ""){
  x = as.vector(x)
  x.ab = x[x > 0]
  x.ab.ranked = x.ab[order(x.ab, decreasing = TRUE)]
  return(x.ab.ranked)
}
```

Now, let's examine the RAC for `site1` of the BCI data set.

In the R code chunk below, do the following:

1. Create a sequence of ranks and plot the RAC with natural-log-transformed abundances,

2. Label the x-axis "Rank in abundance" and the y-axis "log(abundance)"

```{r}
plot.new()
site1 = BCI[1, ]

rac = RAC(x = site1)
ranks = as.vector(seq(1, length(rac)))
opar = par(no.readonly = TRUE)                         # saves default plot parameters
par(mar = c(5.1, 5.1, 4.1, 2.1))                       # new settings for par
plot(ranks, log(rac), type = 'p', axes = F,            # plots w/o axes
     xlab = "Rank in abundance", ylab = "Abundance", 
     las = 1, cex.lab = 1.4, cex.axis = 1.25)

box()                                                  # manually adds border
axis(side = 1, labels = T, cex.axis = 1.25)            # manually adds x-axis
axis(side = 2, las = 1, cex.axis = 1.25,               # manually adds y-axis
     labels = c(1, 2, 5, 10, 20), at = log(c(1, 2, 5, 10, 20)))

par = opar                                             # reset plotting parameters
```

***Question 5***: What effect does visualizing species abundance data on a log-scaled axis have on how we interpret evenness in the RAC?

> ***Answer 5***: In general, sites have a few abundant taxa and many rare taxa; log-transforming the abundance data gives a linear shape on most ecological data.


Now that we have visualized unevennes, it is time to quantify it using Simpson's evenness ($E_{1/D}$) and Smith and Wilson's evenness index ($E_{var}$).

### Simpson's evenness ($E_{1/D}$)

In the R code chunk below, do the following:

1. Write the function to calculate $E_{1/D}$, and

2. Calculate $E_{1/D}$ for `site1`.

```{r}
SimpE = function(x = ""){
  S = S.obs(x)
  x = as.data.frame(x)
  D = diversity(x, "inv")
  E = (D)/S
  return(E)
}

site1 = BCI[1, ]
SimpE(site1)
```

### Smith and Wilson's evenness index ($E_{var}$)

In the R code chunk below, please do the following:

1. Write the function to calculate $E_{var}$,

2. Calculate $E_{var}$ for `site1`, and

3. Compare $E_{1/D}$ and $E_{var}$.

```{r}
Evar = function(x){
  x = as.vector(x[x > 0])
  1- (2/pi) * atan(var(log(x)))
}

Evar(site1)
SimpE(site1)
```

***Question 6***: Compare estimates of evenness for `site1` of BCI using $E_{1/D}$ and $E_{var}$.
Do they agree? If so, why? If not, why? What can you infer from the results.

> ***Answer 6***: They are somewhat similar, but they are not the same; the Smith and Wilson's index shows higher eveness than Simpson's, because it is calculated to decrease bias towards more abundant species.


##5) INTEGRATING RICHNESS AND EVENNESS: DIVERSITY METRICS

So far, we have introduced two primary aspects of diversity, i.e., richness and evenness.
Here, we will use popular indices to estimate diversity, which explicitly incorporate richness and evenness
We will write our own diversity functions and compare them against the functions in `vegan`. 

### Shannon's diversity (a.k.a., Shannon's entropy)

In the R code chunk below, please do the following:

1. Provide the code for calculating H' (Shannon's diversity),

2. Compare this estimate with the output of `vegan`'s diversity function using method = "shannon".

```{r}
ShanH = function(x = ""){
  H = 0
  for (n_i in x){
    if(n_i > 0){
      p = n_i / sum(x)
      H = H - p*log(p)
    }
  }
  return(H)
}

ShanH(site1)
diversity(site1, index = "shannon")
```

### Simpson's diversity (or dominance)

In the R code chunk below, please do the following:

1. Provide the code for calculating D (Simpson's diversity),

2. Calculate both the inverse (1/D) and 1 - D,

2. Compare this estimate with the output of `vegan's` diversity function using method = "simp".

```{r}
SimpD = function(x = ""){
  D = 0
  N = sum(x)
  for (n_i in x){
    D = D + (n_i^2)/(N^2)
  }
  return(D)
}

D.inv = 1/SimpD(site1)
D.sub = 1-SimpD(site1)

D.inv
D.sub

diversity(site1, "inv")
diversity(site1, "simp")
```

***Question 7***: Compare estimates of evenness for `site1` of BCI using $E_{H'}$ and $E_{var}$.
Do they agree? If so, why? If not, why? What can you infer from the results.

> ***Answer 7***: I'm not sure how to compare these, since $E_{var}$ is measuring evenness and $E_{H'}$ is measuring diversity, which combines evenness and richness. 

### Fisher's $\boldsymbol\alpha$

In the R code chunk below, please do the following:

1. Provide the code for calculating Fisher's $\boldsymbol\alpha$,

2. Calculate Fisher's $\boldsymbol\alpha$ for `site1` of BCI.

```{r}
rac = as.vector(site1[site1 > 0])
invD = diversity(rac, "inv")
invD

Fisher = fisher.alpha(rac)
Fisher
```

***Question 8***: How is Fisher's $\boldsymbol\alpha$ different from $E_{H'}$ and $E_{var}$? What does Fisher's $\boldsymbol\alpha$ take into account that $E_{H'}$ and $E_{var}$ do not?

> ***Answer 8***: Fisher's alpha accounts for richness, which both Simpson's and Smith and Wilson's measures of evenness do not.

##6) MOVING BEYOND UNIVARIATE METRICS OF $\alpha$ DIVERSITY

The diversity metrics that we just learned about attempt to integrate richness and evenness into a single, univariate metric.
Although useful, information is invariably lost in this process. 
If we go back to the rank-abundance curve, we can retrieve additional information -- and in some cases -- make inferences about the processes influencing the structure of an ecological system.

## Species abundance models

The RAC is a simple data structure that is both a vector of abundances.
It is also a row in the site-by-species matrix (minus the zeros, i.e., absences). 

Predicting the form of the RAC is the first test that any biodiversity theory must pass and there are no less than 20 models that have attempted to explain the uneven form of the RAC across ecological systems.

In the R code chunk below, please do the following:

1. Use the `radfit()` function in the `vegan` package to fit the predictions of various species abundance models to the RAC of `site1` in BCI,

2. Display the results of the `radfit()` function, and 

3. Plot the results of the `radfit()` function using the code provided in the handout.

```{r}
RACresults = radfit(site1)
RACresults

plot.new()
plot(RACresults, las = 1, cex.lab = 1.4, cex.axis = 1.25)
```


***Question 9***: Answer the following questions about the rank abundance curves:
a) Based on the output of `radfit()` and plotting above, discuss which model best fits our rank-abundance curve for `site1`?
b) Can we make any inferences about the forces, processes, and/or mechanisms influencing the structure of our system, e.g., an ecological community?

> ***Answer 9a***:  Mandelbrot
> ***Answer 9b***: We can infer that there ARE forces/mechanisms acting to structure this system, because the data was not well-described by the null model-- but without knowing more, we cannot make inferences about what those forces/mechanisms actually are.


***Question 10***: Answer the following questions about the preemption model:
a.  What does the preemption model assume about the relationship between total abundance (*N*) and total resources that can be preempted?
b.  Why does the niche preemption model look like a straight line in the RAD plot?

> ***Answer 10a***:  It assumes that N is determined only by total resources, and that each additional species partitions the resources in a predictable and consistent way. To be honest, I do not entirely understand this, even after reading a bit about it.
> ***Answer 10b***: It has been log-transformed, so that the constant rate of decay appears to be a constant rate of decrease



***Question 11***: Why is it important to account for the number of parameters a model uses when judging how well it explains a given set of data?

> ***Answer 11***: You can get any model to fit any data if you add enough parameters, so a large number of parameters could mean that the model has been overworked and is only fitting artificially, rather than being a good explanation for the shape of the data.


## SYNTHESIS

1.  As stated by Magurran (2004) the  ${D = } \sum p_i^2$ derivation of Simpson's Diversity only applies to communities of infinite size. For anything but an infinitely large community, Simpson's Diversity index is calculated as ${D = } \sum \frac{n_i(n_i -1)} {N(N-1)}$. 
Assuming a finite community, calculate Simpson's D, 1 - D, and Simpson's inverse (i.e. 1/D) for `site 1` of the BCI site-by-species matrix. 

```{r}
SimpDf = function(x = ""){
  D = 0
  N = sum(x)
  for (n_i in x){
    D = D + (n_i*(n_i-1)) / (N*(N-1))
  }
  return(D)
}

SimpDf(site1)
Df.inv = 1/SimpDf(site1)
Df.sub = 1-SimpDf(site1)
```


2.  Along with the rank-abundance curve (RAC), another way to visualize the distribution of abundance among species is with a histogram (a.k.a., frequency distribution) that shows the frequency of different abundance classes.
For example, in a given sample, there may be 10 species represented by a single individual, 8 species with two individuals, 4 species with three individuals, and so on.
In fact, the rank-abundance curve and the frequency distribution are the two most common ways to visualize the species-abundance distribution (SAD) and to test species abundance models and biodiversity theories.
To address this homework question, use the R function **hist()** to plot the frequency distribution for `site 1` of the BCI site-by-species matrix, and describe the general pattern you see.

```{r}
site1.vec = as.vector(site1[1, ], mode = "numeric")
hist(site1.vec, breaks = seq(0, 25, by=1))
```

> There are many, many species that are not represented by any individuals. There are fewer species with 1-5 individuals, then only a handful of species with 6 or more individuals. The dramatic dropoff between the number of species with 0 and 1 individuals representing them is a good demonstration of why this kind of data is often better visualized when log-transformed; this histogram is a great visualization of actual species distributions, but if you want to be able to read more than the number of unrepresented species with any accuracy, a log-transformed graph will be more useful.

3.  We asked you to find a biodiversity dataset with your partner.
This data could be one of your own or it could be something that you obtained from the literature. 
Load that dataset.
How many sites are there?
How many species are there in the entire site-by-species matrix?
Any other interesting observations based on what you learned this week?

```{r}
SpeciesbySite = read.table("~/GitHub/QB2019_Caple/data_from_brianna.csv", header = TRUE, sep = ",")
species = SpeciesbySite[ ,1]
SitebySpecies = as.data.frame(t(SpeciesbySite[ ,-1]))
colnames(SitebySpecies) = species

Site.richness = S.obs(SitebySpecies)
summary(Site.richness)
```

> There are 20 sites and 302 species.
> This is a presence/absence dataset, so I will not be able to compute scores for coverage or evenness. The "S.obs" function that we wrote shows that the species richness of the sites ranges from 21 to 97 species, with a mean of 47.75 species. 

## SUBMITTING YOUR ASSIGNMENT
Use Knitr to create a PDF of your completed alpha_assignment.Rmd document, push it to GitHub, and create a pull request.
Please make sure your updated repo include both the HTML and RMarkdown files.

Unless otherwise noted, this assignment is due on **Wednesday, January 23^rd^, 2017 at 12:00 PM (noon)**.