---
title: "7. Control Structures"
output: html_document
---

1) Introduction to Control Structures

A. 'If, Else' Logical Tests

```{r}
x = sample(1:15, size = 1)

if(x <= 10) {
  paste("x is", x, "and is less than, or equal to, 10")
} else{
  paste("x is", x, "and is greater than 10")
}

x = sample(1:15, size = 1)

if(x==10) {
  paste("x is", x, "and is equal to 10")
} else if(x < 10) {
  paste("x is", x, "and is less than 10")
} else{
  paste("x is", x, "and is greater than 10")
}
```

B. For-Loops

```{r}
for(i in 1:10) {
  print(i)
}

x = c("apples", "oranges", "bananas", "strawberries")  # x is a vector of fruit names

for(i in x) {
  print(i)
}

# a for-loop can iterate over any type of vector
for(i in seq(x)) {
  print(x[i])
}

# for-loops can also transform vectors
for(i in seq(x)) {
  len = nchar(x[i])
}

len = ""
for(i in seq(x)) {
  len[i] = nchar(x[i])
}
print(len)

# can use 'apply' functions for many of the same uses; they are usually much faster
len = sapply(x, nchar)
# the output of saaply is a "named int", but this can easily be coerced to a vector
len = as.vector(len)
print(len)
```

C. Combining control structures: 'for' and 'if'

```{r}
for(i in 1:10) {
  if(i %% 3 == 0) {
    print(i)
  }
}
```

D. While loops

```{r}
i = 1
while(i < 10) {
  print(i)
  i = i + 1
}
```


2) Applying Control Structures to Diversity Data

B. Generating the Source Community

```{r}
# clear the environment
rm(list=ls())

# for reproducible results we will set the seed
set.seed(6)

# draw 20 random numbers from log-normal distribution with a mean of 5 (log scale)
jelly.jar = rlnorm(n = 20, meanlog = 5)

# force numbers to integers
jelly.jar = as.integer(jelly.jar)

# generate and plot the rank-abundance curve
require(vegan)
RACresults = rad.lognormal(jelly.jar)
plot(RACresults, las = 1, cex.lab = 1.4, cex.axis = 1.25)
```

B. Sampling the Source Community

```{r}
sample.community = function(x,n){
  # write out an explicit vector of the community
  all.individuals = rep(seq(x), x)
  
  # take n samples from the community
  survey = sample(all.individuals, n)
  
  # prepare vector of species with 0 in each bin
  survey.sum = rep(0, length(x))
  
  # add 1 to species bin each time an individual of that species was sampled
  for(i in survey){
    survey.sum[i] = survey.sum[i] + 1
  }
  return(survey.sum)
}

# set seed
set.seed(3)

# sample using our function
sample.jelly = sample.community(x = jelly.jar, n = 100)

# generate and plot RAC using vegan
RACsample = rad.lognormal(sample.jelly)
plot.new()
par(mfrow = c(1, 2))
plot(RACresults, las = 1, cex.lab = 1.4, cex.axis = 1.25, xlim = c(1, 20), main = "Source community")
plot(RACsample, las = 1, cex.lab = 1.4, cex.axis = 1.25, xlim = c(1, 20), main = "Sample")
```

Question: how similar or different is the sample from the source community?

Testing input to a function using the if statement

```{r}
# breaking a function by attempting to sample with a larger n than the community
# sample.community(x = jelly.jar, n = (sum(jelly.jar)+1))

# rewrite function with error prevention
sample.community = function(x,n){
  
  # sample size cannot exceed community size
  if(n > sum(x)) stop ("Sample size cannot exceed total size of community"
                       )
  # write out an explicit vector of the community
  all.individuals = rep(seq(x), x)
  
  # take n samples from the community
  survey = sample(all.individuals, n)
  
  # prepare vector of species with 0 in each bin
  survey.sum = rep(0, length(x))
  
  # add 1 to species bin each time an individual of that species was sampled
  for(i in survey){
    survey.sum[i] = survey.sum[i] + 1
  }
  return(survey.sum)
}

sample.community(x = jelly.jar, n = (sum(jelly.jar)+1))
```

C. Estimating Richness via Resampling

```{r}
S = estimateR(sample.jelly)
S

# looking at how different indices estimate the true diversity

# number of individuals per sample
n = 100
# number of repeated samples
N = 1000
# initialize matrix to store results
repeated.S = matrix(NA, nrow = N, ncol = length(S))
colnames(repeated.S) = names(S)

# resample and store results
for(i in seq(N)){
  sample.jelly = sample.community(x = jelly.jar, n = n)
  repeated.S[i,] = as.vector(estimateR(sample.jelly))
}

# plot distribution of results as histogram
plot.new()
par(mfrow = c(1, 3))
# find range of all data so plots have the same x-axis
x.lim = range(repeated.S[ , c("S.obs", "S.chao1", "S.ACE")])

for(i in c("S.obs", "S.chao1", "S.ACE")){
  hist(repeated.S[,i], prob = TRUE, cex.lab = 1.4, cex.axis = 1.25,
       xlab = i, main = NULL, col = "grey", las = 1,
       xlim = x.lim, breaks = seq(x.lim[1], x.lim[2]+1, by = 2))
  # add dashed line to mark true S
  abline(v = length(jelly.jar), lty = 2, lwd = 3)
}
```

E. Collector's Curve

```{r}
# take a single sample of size N from the source community
N <- 1000
sample.jelly <- sample.community(jelly.jar, n = N)

# generate vector to store results
S.collectors <- rep(NA, N)

# loop through 1:N and check S.obs for sub-samples
for(i in 1:N){
  sub.sample <- sample.community(sample.jelly, n = i)
  S.collectors[i] <- estimateR(sub.sample)["S.obs"]
}

# plot results
plot.new()
par(mfrow = c(1, 1))
plot(1:N, S.collectors, xlab = "Sample size", ylab = "S.obs")
```

Question: Given limited resources (time and money), what sample size would you use for estimating richness of the source community?

E. Shannon's Diversity: A Mathematical for Loop

```{r}
ShanH <- function(x){
  H = 0
  for(i in seq(x)){
    p_i = x[i] / sum(x)
    H = H + p_i * log(p_i)
  }
  return(-H)
}

ShanH(jelly.jar)
```